The Folly of Passwords

### 2014-12-12 paper.tex
Fair progress on chapter 3 now.

Ultimately, if we can evaluate the use of each alternative as if it were
to be used as a single-factor, it should not prevent an organisation
adopting the alternative alongside password authentication in principle. We
will give some consideration for the ease of doing so in our evaluation. We
would advise always doing so as part of a design that considers the users,
models their use cases and ensures that the approach is indeed increasing
security more than it is just moving the costs onto the users.

@@

>>>>

Ultimately, if we can evaluate the use of each alternative as if it were
to be used as a single-factor, it should not prevent an organisation
adopting the alternative alongside password authentication in principle. We
gave some consideration for the ease of doing so in our evaluation. We also
would advise always doing so as part of a design that considers the users,
models their use cases and ensures that the approach is indeed increasing
security more than it is just moving the costs onto the users.

@@

The chosen OTP system was the OATH Toolkit and specifically its
module for the Unix Pluggable Authentication Module (PAM) system
\parencite{samar1996unified}.

The OATH PAM module was simple to configure on a Debian GNU/Linux desktop
computer and provided one-time password login using any OTP token or mobile
app, such as Google Authenticator. Our setup was tested with both the
command line tool from the OATH Toolkit as well as with the
Google Authenticator Android app.

The challenge here is that each user needs a unique secret which has to
be managed on a trusted server. The secret for every employee then needs
to be available in a configuration file on every desktop computer. This
can be achieved with configuration management systems such as Puppet,
Chef, Ansible, Salt, etc.

Another option is to try to centralise the OTP authentication.
Solutions such as LinOTP suggest the abiity to have a centralised
OTP-based authentication system with a custom PAM module. Combined
with Kerberos, true Single Sign-On should then be possible
including for access to email and web-based applications.

A centralised solution is more flexible for corporate systems
as PAM does not provide an intrinsic method for fowarding local
authentication on to other services. Microsoft Windows achieved
this in the past with NTLM -- which is possible to configure
also on Unix systems with Winbind -- but that is largely
regarded as an insecure mechanism today.

Some desktop environments (e.g. Gnome, Mac OS X, etc.) provide
a local keychain that can store SSH keys, SSL certificates, etc.
which are unlocked whilst the user is logged in, but inaccessible
otherwise. This already provides a convenience desktop
users, but it relies on the user authenticating with a password
that is then the same password used to update the keychain
(Mac OS X updates the keychain password when the user's main
account password is changed). There maybe be some scope for
further research into whether dekstop keychains can be made
compatible with OTP authentication.

There is certainly scope for further work to evaluate extensive
use of LinOTP and Kerberos within an enterprise setting and the
benefits they bring over relying on a local desktop keychain.

The PAM USB module for the PAM system is similar in many
ways to the OATH OTP module in terms of setup. The module
installs easily on Debian GNU/Linux systems and it provides
tools to manage a single configuration file that stores a
mapping between users and an identifier for their USB device.

The module also provides a mechanism for locking computers
when the USB key is removed, which is helpful in corporate
environments where users are instructed to lock their desktops
when they are unattended. This mechanism is in line with
drivers removing their keys when they leave their car
unattended, so this is more in line with existing user
habits than asking users not to disclose or to strengthen
their passwords.

However, this PAM module has the same limitations as the
OATH OTP module in that the very simple local authenication
is not easily forwarded to other services on the network.

An advantage this module has over the OATH OTP module is that
local applications can make authentication requests
multiple times without the user needing to be involved. As
long as the USB device remains connected, applications
can authenticate multiple times, but OTP authentication
will require a new password to be entered on each authentication
request.
